---
title: "dearseq_adjustDay0"
author: "Madeleine"
date: "2023-11-03"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("dearseq")
load("metadata.rdata")
sampleIDs <- metadata$SampleID
rawcounts <- read.csv("DM_raw-counts_transposed.csv") 
genes <- rawcounts$Genes
rownames(rawcounts) <- genes

#rename columns to match metadata
name1 <- colnames(rawcounts[, -1])
name1 <- chartr(".", "-", name1)
name1 <- gsub("X", "", name1)
colnames(rawcounts) <- c("Genes", name1)
dim(rawcounts)

#select columns that are in our metadata
rawcounts <- select(rawcounts, c("Genes", all_of(sampleIDs)))
dim(rawcounts)
#removes 13 columns, now have 72 samples (4 for each participant) and 1 column with gene name
```

Remove genes with 0 observations in all samples (n=6,289)
```{r}
#remove genes with 0 observations in all samples
gene_no0 <- rowSums(select(rawcounts, -1)) >0
rawcounts_no0 <- filter(rawcounts, gene_no0)
dim(rawcounts_no0)
#removes 6,289 genes with all zeros
```

Normalizing the day 1 genes
```{r norm1}
library(edgeR)
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genes_normfact <- edgeR::calcNormFactors(genesDGE, method = "TMM")
normfact_no0_cpm2p_TMMlog2cpm <- edgeR::cpm(genes_normfact, log = T)
```

Creating the design matrix for normalized day 1 genes
```{r design1norm}
genesDay1_normfact <- genes_normfact[,metadata$VisitID==1]
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay1_normfact$samples))
samplesDay1 <- metadata %>% 
  filter(VisitID ==1)
samplesDay1 <- samplesDay1$SampleID
design_day1 <- design %>% filter(row.names(design) %in% samplesDay1)
design_day1 <- design_day1 %>% 
  as.matrix()
```

Running dearSeq on day 1 normalized genes
```{r dearseq1norm}
res_dearseq1_a <- dearseq::dear_seq(exprmat=genesDay1_normfact$counts, variables2test = design_day1[,"LogPRNT50_Day28", drop = F], covariates = design_day1[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", preprocessed = T)
summary(res_dearseq1_a)
```

Leaving genes non-normalized for dearseq to normalize
```{r desgin1}
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genesDay1 <- genesDGE[, metadata$VisitID==1]
```

Creating the design matrix of non-normalized day 1 genes
```{r}
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay1$samples))
samplesDay1 <- metadata %>% 
  filter(VisitID ==1)
samplesDay1 <- samplesDay1$SampleID
design_day1 <- design %>% filter(row.names(design) %in% samplesDay1)
design_day1 <- design_day1 %>% 
  as.matrix()
```

Running dearseq on day 1 non-normalized genes
```{r dearseq1}
res_dearseq1_b <- dearseq::dear_seq(exprmat=genesDay1$counts, variables2test = design_day1[,"LogPRNT50_Day28", drop = F], covariates = design_day1[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", preprocessed = F)
summary(res_dearseq1_b)
```

### Day 3
Taking day 3 normalized genes
```{r norm3}
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genes_normfact <- edgeR::calcNormFactors(genesDGE, method = "TMM")
genesDay3_normfact <- genes_normfact[,metadata$VisitID==3]
```

Making the design matrix with day 3 normalized genes
```{r design3norm}
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay3_normfact$samples))
samplesDay3 <- metadata %>% 
  filter(VisitID ==3)
samplesDay3 <- samplesDay3$SampleID
design_day3 <- design %>% filter(row.names(design) %in% samplesDay3)
design_day3 <- design_day3 %>% 
  as.matrix()
```

Running dearseq for day 3 normalized genes
```{r}
res_dearseq3_a <- dearseq::dear_seq(exprmat=genesDay3_normfact$counts, variables2test = design_day3[,"LogPRNT50_Day28", drop = F], covariates = design_day3[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", preprocessed = T)
summary(res_dearseq3_a)
plot(res_dearseq3_a)
#0 genes
```

Repeating on non-normalized day 3 genes
```{r}
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genesDay3 <- genesDGE[, metadata$VisitID==3]
```

Creating design matrix for day 3 genes non-normalized
```{r}
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay3$samples))
samplesDay3 <- metadata %>% 
  filter(VisitID ==3)
samplesDay3 <- samplesDay3$SampleID
design_day3 <- design %>% filter(row.names(design) %in% samplesDay3)
design_day3 <- design_day3 %>% 
  as.matrix()
```

Running dearseq for day 3 non-normalized genes
```{r }
res_dearseq3_b <- dearseq::dear_seq(exprmat=genesDay3$counts, variables2test = design_day3[,"LogPRNT50_Day28", drop = F], covariates = design_day3[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", preprocessed = F)
summary(res_dearseq3_b)
plot(res_dearseq3_b)
#0 genes
```


### Day 1 and 3 combined
taking normalized genes on day 1 and 3
```{r}
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genes_normfact <- edgeR::calcNormFactors(genesDGE, method = "TMM")
genesDay13_normfact <- genes_normfact[,metadata$VisitID %in% c(1, 3)]
```

Creating the design matrix for normalized day 1 and day 3 genes
```{r}
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay13_normfact$samples))
samplesDay13 <- metadata %>% 
  filter(VisitID ==3 | VisitID == 1)
samplesDay13 <- samplesDay13$SampleID
design_day13 <- design %>% filter(row.names(design) %in% samplesDay13)
design_day13 <- design_day13 %>% 
  as.matrix()
```

Running dearseq on day 1 and 3 normalized genes
```{r}
res_dearseq13_a <- dearseq::dear_seq(exprmat=genesDay13_normfact$counts, variables2test = design_day13[,"LogPRNT50_Day28", drop = F], covariates = design_day13[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", sample_group = rep(1:18, each = 2), preprocessed = T)
summary(res_dearseq13_a)
plot(res_dearseq13_a)
#0 genes
```

Subsetting the non-normalized genes
```{r}
genesDGE <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genesDay13 <- genesDGE[, metadata$VisitID %in% c(1,3)]
```

creating the design matrix of non-normalized day 1 and day 3 genes
```{r}
design <- as.data.frame(model.matrix(~LogPRNT50_Day28 + LogPRNT50_Day0, data=genesDay13$samples))
samplesDay13 <- metadata %>% 
  filter(VisitID ==3 | VisitID==1)
samplesDay13 <- samplesDay13$SampleID
design_day13 <- design %>% filter(row.names(design) %in% samplesDay13)
design_day13 <- design_day13 %>% 
  as.matrix()
```

```{r}
res_dearseq13_b <- dearseq::dear_seq(exprmat=genesDay13$counts, variables2test = design_day13[,"LogPRNT50_Day28", drop = F], covariates = design_day13[,c("(Intercept)", "LogPRNT50_Day0"), drop = F], which_test = "permutation", sample_group = rep(1:18, each = 2), preprocessed = F)
summary(res_dearseq13_b)
plot(res_dearseq13_b)
#0 genes
```


