---
title: "dearseq_logfold"
author: "Madeleine"
date: "2023-10-23"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(limma)
setwd("~/Documents/OMICS/Project")
load("cleandata.rdata")
sampleIDs <- metadata$sampleID
rawcounts <- read_excel("Raw-counts_transposed Results.xlsx") %>% 
  select(c( "Genes", all_of(sampleIDs)))
dim(rawcounts)
```

### Day 1 with filtering 
**Filtering** 
```{r}
gene_no0 <- rowSums(select(rawcounts, -1)) >0
rawcounts_no0 <- filter(rawcounts, gene_no0)
dim(rawcounts_no0)
mycpmfun <- function(x) { 
  return((x/sum(x))*10^6)
}

cpm <-apply(select(rawcounts_no0, -1), MARGIN = 2, FUN = mycpmfun)
rawcounts_no0_cpm2p <- filter(rawcounts_no0, rowSums(cpm>2)>=5)
dim(rawcounts_no0_cpm2p)
```

**Normalization**
```{r}
library(edgeR)
genesDGE <- DGEList(counts = select(rawcounts_no0_cpm2p, -1), samples = metadata, genes = select(rawcounts_no0_cpm2p, 1))
genes_normfact_f <- edgeR::calcNormFactors(genesDGE, method = "TMM")
#rawcounts_no0_cpm2p_TMMlog2cpm <- edgeR::cpm(genes_normfact, log = T)
```

Design matrix
```{r}
genesDay1_normfact_f <- genes_normfact_f[,metadata$VisitID==1]
design <- as.data.frame(model.matrix(~logFold, data=genes_normfact_f$samples))
samplesDay1 <- metadata %>% 
  filter(VisitID ==1)
samplesDay1 <- samplesDay1$sampleID
design_day <- design %>% filter(row.names(design) %in% samplesDay1)
design_day1 <- design_day1 %>% 
  as.matrix()
```

**Edge R Pipeline**
```{r}
genesDay1_normfact_f_commondisp <- edgeR::estimateCommonDisp(genesDay1_normfact_f)
fit_1_f <- glmFit(y=genesDay1_normfact_f_commondisp, design = design_day1)
lrt_1_f <- glmLRT(glmfit = fit_1_f, coef=2)
signif_edgeR <- decideTestsDGE(lrt_1_f, adjust.method = "BH", p.value = 0.05, lfc = 1)
summary(signif_edgeR)
genelist_edgeR <- genesDay1_normfact_f_commondisp$genes$Genes[as.logical(signif_edgeR)]
```

```{r}
v_1_f <- voom(counts = genesDay1_normfact_f, design = design_day1, plot = T)
fit_1_f <- lmFit(v_1_f, design = design_day1)
fit_1_f <- eBayes(fit_1_f, robust = T)
voomlimma_signif <- decideTests(fit_1_f, adjust.method = "BH", p.value = 0.05, lfc=1)
summary(voomlimma_signif)
genelist_voomlimma_f <- genesDay1_normfact_f$genes$Genes[as.logical(voomlimma_signif)]
```

Volcano Plot - this isn't working
```{r}
adj_pval <- p.adjust(fit_1_f$p.value[,2], method = "BH")
logfoldchange_1_f <- fit_1_f$coefficients[,2]
signif <- ifelse(adj_pval < 0.05 & logfoldchange_1_f>1, "Up-regulated", "Non-significant")
signif[logfoldchange_1_f<-1 & adj_pval <0.05] <- "Down-regulated"
volcano_data <- cbind.data.frame(ml10_Adjusted_Pval = -log10(adj_pval), Log2_FoldChange = logfoldchange_1_f, Signif = signif, Gene = genesDay1_normfact_f$genes$Genes)

ggplot(volcano_data) + 
  geom_point(aes(x=Log2_FoldChange, 
                 y = ml10_Adjusted_Pval, 
                 col = signif), alpha = 0.2) +
  geom_hline(aes(yintercept = -log10(0.05)), linetype = 2, color = "grey75") +
  geom_vline(aes(xintercept = log2(1/2)), linetype = 2, color = "blue") + 
  geom_vline(aes(xintercept = log2(2)), linetype = 2, color = "red") + 
  annotation_logticks(sides = "l") + 
  scale_y_continuous(breaks = c(0:3)*2, labels = paste0("1e-0", c(0:3)*2)) + 
  scale_x_continuous(breaks = log2(c(1/8, 1/2, 1, 2, 8)), labels = c("1/8", "1/2", "1", "2", "8"), limits =log2(c(1/16, 16))) + 
  scale_color_manual(name = "Significance", values = c("blue", "grey75", "red")) + 
  theme_bw() + 
  ggrepel::geom_label_repel(data = filter(volcano_data, abs(Log2_FoldChange)>log2(4) & ml10_Adjusted_Pval > -log10(1e-04)), aes(x=Log2_FoldChange, y = ml10_Adjusted_Pval, label = Gene))
```

DearSeq
```{r}
res_dearseq1_f_a <- dearseq::dear_seq(exprmat=genesDay1_normfact_f$counts, variables2test = design_day1[,"logFold", drop = F], covariates = design_day1[,"(Intercept)", drop = F], which_test = "asymptotic", preprocessed = F)
summary(res_dearseq1_f_a)
set.seed(100)
res_dearseq1_f_p <- dearseq::dear_seq(exprmat=genesDay1_normfact_f$counts, variables2test = design_day1[,"logFold", drop = F], covariates = design_day1[,"(Intercept)", drop = F], which_test = "permutation", preprocessed =F)
summary(res_dearseq1_f_p)
# no significant genes when filtered
```

### Day 1 without filtering 
Filtering 
```{r}
genesDGE_nf <- DGEList(counts = select(rawcounts_no0, -1), samples = metadata, genes = select(rawcounts_no0, 1))
genes_normfact_nf <- edgeR::calcNormFactors(genesDGE_nf, method = "TMM")
#rawcounts_no0_TMMlog2cpm_nf <- edgeR::cpm(genes_normfact_nf, log = T)
```

Design matrix
```{r}
genesDay1_normfact_nf <- genes_normfact_nf[,metadata$VisitID==1]
design <- as.data.frame(model.matrix(~logFold, data=genes_normfact_nf$samples))
samplesDay1 <- metadata %>% 
  filter(VisitID ==1)
samplesDay1 <- samplesDay1$sampleID
design <- design %>% filter(row.names(design) %in% samplesDay1)
design_day1 <- design %>% 
  as.matrix()
```

**Edge R Pipeline**
```{r}
genesDay1_normfact_nf_commondisp <- edgeR::estimateCommonDisp(genesDay1_normfact_nf)
fit_1_nf <- glmFit(y=genesDay1_normfact_nf_commondisp, design = design_day1)
lrt_1_nf <- glmLRT(glmfit = fit_1_nf, coef=2)
signif_edgeR <- decideTestsDGE(lrt_1_nf, adjust.method = "BH", p.value = 0.05, lfc = 1)
summary(signif_edgeR)
genelist_edgeR <- genesDay1_normfact_nf_commondisp$genes$Genes[as.logical(signif_edgeR)]
```

Voom-limma pipeline
```{r}
v_1_nf <- voom(counts = genesDay1_normfact_nf, design = design_day1, plot = F)
fit_1_nf <- lmFit(v_1_nf, design = design_day1)
logfoldchange_1_nf <- fit_1_nf$coefficients[,2]

fit_1_nf <- eBayes(fit_1_nf, robust = T)
voomlimma_signif <- decideTests(fit_1_nf, adjust.method = "BH", p.value = 0.05, lfc=1)
genelist_voomlimma_nf <- genesDay1_normfact_nf$genes$Genes[as.logical(voomlimma_signif)]
summary(voomlimma_signif)
```

DearSeq
```{r}
res_dearseq1_nf_a <- dearseq::dear_seq(exprmat=genesDay1_normfact_nf$counts, variables2test = design_day1[,"logFold", drop = F], covariates = design_day1[,"(Intercept)", drop = F], which_test = "asymptotic", preprocessed = F)
summary(res_dearseq1_nf_a)
# 0 genes
set.seed(100)
res_dearseq1_nf_p <- dearseq::dear_seq(exprmat=genesDay1_normfact_nf$counts, variables2test = design_day1[,"logFold", drop = F], covariates = design_day1[,"(Intercept)", drop = F], which_test = "permutation", preprocessed = F)
#removes 2072 genes with 0 variance from testing procedure, "should've been removed beforehand"
summary(res_dearseq1_nf_p)
plot(res_dearseq1_nf_p)
# 7 significant genes when not filtering 

dearseq_signif_1_nf_p <- res_dearseq1_nf_p$pvals$adjPval<0.05&logfoldchange_1_nf >1
geneList_1_nf_p <- genesDay1_normfact_nf$genes$Genes[as.logical(dearseq_signif_1_nf_p)]
```

### Day 3 with filtering 
Design matrix
```{r}
genesDay3_normfact_f <- genes_normfact_f[,metadata$VisitID==3]
design <- as.data.frame(model.matrix(~logFold, data=genes_normfact_f$samples))
samplesDay3 <- metadata %>% 
  filter(VisitID ==3)
samplesDay3 <- samplesDay3$sampleID
design_day3 <- design %>% filter(row.names(design) %in% samplesDay3)
design_day3 <- design_day3 %>% 
  as.matrix()
```

**Edge R Pipeline**
```{r}
genesDay3_normfact_f_commondisp <- edgeR::estimateCommonDisp(genesDay3_normfact_f)
fit_3_f <- glmFit(y=genesDay3_normfact_f_commondisp, design = design_day3)
lrt_3_f <- glmLRT(glmfit = fit_3_f, coef=2)
signif_edgeR <- decideTestsDGE(lrt_3_f, adjust.method = "BH", p.value = 0.05, lfc = 1)
summary(signif_edgeR)
genelist_edgeR <- genesDay3_normfact_f_commondisp$genes$Genes[as.logical(signif_edgeR)]
```

Voom-limma pipeline
```{r}
v_3_f <- voom(counts = genesDay3_normfact_f, design = design_day3, plot = F)
fit_3_f <- lmFit(v_3_f, design = design_day3)
logfoldchange_3_f <- fit_3_f$coefficients[,2]

fit_3_f <- eBayes(fit_3_f, robust = T)
voomlimma_signif <- decideTests(fit_3_f, adjust.method = "BH", p.value = 0.05, lfc=1)
genelist_voomlimma_f <- genesDay3_normfact_f$genes$Genes[as.logical(voomlimma_signif)]
summary(voomlimma_signif)
```

DearSeq
```{r}
res_dearseq3_f_a <- dearseq::dear_seq(exprmat=genesDay3_normfact_f$counts, variables2test = design_day3[,"logFold", drop = F], covariates = design_day3[,"(Intercept)", drop = F], which_test = "asymptotic", preprocessed = F)
summary(res_dearseq3_f_a)
# 0 genes 
set.seed(100)
res_dearseq3_f_p <- dearseq::dear_seq(exprmat=genesDay3_normfact_f$counts, variables2test = design_day3[,"logFold", drop = F], covariates = design_day3[,"(Intercept)", drop = F], which_test = "permutation", preprocessed = F)
summary(res_dearseq3_f_p)
#0  genes
```

### Day 3 without filtering 
Design matrix
```{r}
genesDay3_normfact_nf <- genes_normfact_nf[,metadata$VisitID==3]
design <- as.data.frame(model.matrix(~logFold, data=genes_normfact_nf$samples))
samplesDay3 <- metadata %>% 
  filter(VisitID ==3)
samplesDay3 <- samplesDay3$sampleID
design_day3 <- design %>% filter(row.names(design) %in% samplesDay3)
design_day3 <- design_day3 %>% 
  as.matrix()
```

**Edge R Pipeline**
```{r}
genesDay3_normfact_nf_commondisp <- edgeR::estimateCommonDisp(genesDay3_normfact_nf)
fit_3_nf <- glmFit(y=genesDay3_normfact_nf_commondisp, design = design_day3)
lrt_3_nf <- glmLRT(glmfit = fit_3_nf, coef=2)
signif_edgeR <- decideTestsDGE(lrt_3_nf, adjust.method = "BH", p.value = 0.05, lfc = 1)
summary(signif_edgeR)
genelist_edgeR <- genesDay3_normfact_nf_commondisp$genes$Genes[as.logical(signif_edgeR)]
```

Voom-limma pipeline
```{r}
v_3_nf <- voom(counts = genesDay3_normfact_nf, design = design_day3, plot = F)
fit_3_nf <- lmFit(v_3_nf, design = design_day3)
logfoldchange_3_nf <- fit_3_nf$coefficients[,2]

fit_3_nf <- eBayes(fit_3_nf, robust = T)
voomlimma_signif <- decideTests(fit_3_nf, adjust.method = "BH", p.value = 0.05, lfc=1)
genelist_voomlimma_nf <- genesDay3_normfact_nf$genes$Genes[as.logical(voomlimma_signif)]
summary(voomlimma_signif)
```

DearSeq
```{r}
res_dearseq3_nf_a <- dearseq::dear_seq(exprmat=genesDay3_normfact_nf$counts, variables2test = design_day3[,"logFold", drop = F], covariates = design_day3[,"(Intercept)", drop = F], which_test = "asymptotic", preprocessed = F)
summary(res_dearseq3_nf_a)
# 0 genes

set.seed(100)
res_dearseq3_nf_p <- dearseq::dear_seq(exprmat=genesDay3_normfact_nf$counts, variables2test = design_day3[,"logFold", drop = F], covariates = design_day3[,"(Intercept)", drop = F], which_test = "permutation", preprocessed = F)
summary(res_dearseq3_nf_p)
# 7 significant genes

dearseq_signif_3_nf_p <- res_dearseq3_nf_p$pvals$adjPval<0.05&logfoldchange_3_nf >1
geneList_3_nf_p <- genesDay3_normfact_nf$genes$Genes[as.logical(dearseq_signif_3_nf_p)]
```
